Combinatorial Proofs and Decomposition Theorems for First-order Logic

### 2021-01-17 main.tex
still in IX

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w_l}}
\newcommand\wrD {\mathsf{w_r}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\acDeq {\mathsf{ac}^\fequ}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma(#1)}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\left\lfloor{#1}\right\rfloor}

@@

@@

\begin{figure}
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}
            \qquad
            \vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}
            \qquad
            \vlinf{\tttD}{}{\Scons{A\vlan\ttt}}{\Scons A}
            \\ \\[-1ex]
            \vlinf{\sw}{}{\Scons{(A\vlan B)\vlor C}}{\Scons{A\vlan (B\vlor C)}}
            \qquad
            \vlinf{\mix}{}{\Scons{A\vlor B}}{\Scons{A\vlan B}}            
            \\ \\[-1ex]
            \vlinf{\exD}{}{\Scons{\exists x.A}}{\Scons{A\ssubst x t}}
            \qquad
            \vlinf{\fequ}{\text{(where $A\fequ B$)}}{\Scons{A}}{\Scons{B}}
          \end{array}
          $
        }
        \\ \\[-1ex]
        \vlinf{\mfaD}{}{\Scons{\forall x.(A\vlor B)}}{\Scons{(\forall x.A)\vlor(\forall x.B)}}
        \qquad
        \vlinf{\mexD}{}{\Scons{\exists x.(A\vlor B)}}{\Scons{(\exists x.A)\vlor(\exists x.B)}}        
        \\ \\[-1ex]
        \vlinf{\me}{}{\Scons{(A\vlor B)\vlan(C\vlor D)}}{\Scons{(A\vlan C)\vlor(B\vlan D)}}
        \\ \\[-1ex]
        \vlinf{\wrD}{}{\Scons{A\vlor B}}{\Scons{A}}
        \qquad
        \vlinf{\wlD}{}{\Scons{B\vlor A}}{\Scons{A}}
        \qquad
        \vlinf{\acD}{}{\Scons{a}}{\Scons{a\vlor a}}
      \end{array}
      $
    }
  \end{center}
  \caption{Deep inference systems $\FOKS$ (all rules) and $\FOMLS$ (rules in the dashed box)}
  \label{fig:KS1}
\end{figure}

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD}}{}{A}{
        \vlde{\set{\acD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:decompositionA}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there is a formula $A_1$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\cD}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:CP-DI}
  Let $A$ be a rectified formula, and assume we have a combinatorial proof $\phi\colon\gC\to\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'$ such that $\graphof{\rectif{A'}}=\gC$. Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\begin{lemma}
  \label{lem:me}
  The rules $\me,\mfaD,\mexD$ are derivable in $\set{\wlD,\wrD,\cD}$. 
\end{lemma}

@@

\begin{lemma}
  Let $A$ and $B$ be formulas. Then
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\cD,\fequ}}{}{B}{
        \vlhy{A}}}
    \qquad
    \iff
    \qquad
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}
  \end{equation*}
\end{lemma}

@@

%All statements in this section are proved by rule permutation arguments.
\begin{thm}\label{thm:LK1-decompose}
  Let $\Gamma$ be a sequent. If\/ $\sqn\Gamma$ is provable in $\FOLK$ (as depicted on the left below) then there is a sequent
  $\Gamma'$ such that there is a derivation as shown on the right below:\vadjust{\vskip-2ex}
  \begin{equation*}
    \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\Gamma}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}
    \qquad
    \Longrightarrow
    \qquad
    \vlderivation{
      \vlde{\set{\wlD,\wrD,\cD}}{\Deri_2}{\sqns\Gamma}{
        \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\Gamma'}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}}
  \end{equation*}
\end{thm}
\juihsuan{maybe we should take $\form{\Gamma}$ instead? (consider the proof
$\vlderivation{\vlin{\wk}{}{\sqn{\ttt, A}}{
                 \vlin{\ttt}{}{\sqn\ttt}{
                   \vlhy{}}}}$)}

\begin{proof}
  Note that the instances of $\wlD,\wrD,\cD$ in $\Deri_2$ are deep, but inside sequent contexts.
  \todo{}
\end{proof}

@@

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wlD,\wrD}}{}{B}{
        \vlde{\set{\acD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}

@@

In this section we establish the relation between ske bifibrations and derivations  in
$\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wlD,\wrD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices). For a
derivation $\Deri$ in
$\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a rectified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  For $\Deri$, we construct its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances of the rules
  $\rectif\mfaD$ and $\rectif\mexD$ are replaced two steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration. \todo{show the other properties,
    once they are fixed}
\end{proof}
\subsection{From Skew Bifibrations to Contraction and Weakening}

\begin{lemma}\label{skew->cw}
  Let $\phi\colon\gA\to\gB$ be a skew bifibration, and let $A$ and $B$
  be formulas with $\graphof A=\gA$ and $\graphof B=\gB$. Then there
  is a derivation $\vlderivation{
    \vlde{\set{\wlD,\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ with $\mapof\Deri=\phi$.
\end{lemma}

\begin{proof}
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Now let $\PE B_1$ be the propositional formula obtained from
  $\PE B$ by removing all atoms that are not represented by vertices
  in $\vB'$. Then $\graphof{\PE B_1}=\gB_1$. Furthermore, the map
  $\phi'\colon \gB_1\to\gB$ is an injection that reflects
  edges. Hence, by \cite[Proposition~7.6.1]{str:07:RTA}, we
  have a derivation
  $\vlder{\set{\wlD,\wrD}}{\PE\Deri_1}{\PE B}{\PE B_1}$. \todo{show that we have $B_1$ with $\graphof{B_1}=\gB_1$, so that we can can get
  first order derivation $\vlder{\set{\wlD,\wrD}}{\Deri_1}{B}{B_1}$.}

>>>>

\newcommand\aiD {\mathsf{ai}}
\newcommand\faD {\forall}
\newcommand\exD {\exists}
\newcommand\tttD {\ttt}
\renewcommand\wD {\mathsf{w}}
\newcommand\wlD {\mathsf{w}}
\newcommand\wrD {\mathsf{w}}
\renewcommand\cD {\mathsf{c}}
\renewcommand\acD {\mathsf{ac}}
\newcommand\acDeq {\mathsf{ac}^\fequ}
\newcommand\mfaD {\mathsf{m_\forall}}
\newcommand\mexD {\mathsf{m_\exists}}

@@

\newcommand{\rectif}[1]{\widehat{#1}}
\newcommand{\graphof}[1]{\llbracket#1\rrbracket}
\newcommand{\compl}[1]{#1^\complement}
%\newcommand{\sublist}[1]{\set{#1}}
%\newcommand{\subst}[2]{#1\mathord{\scriptstyle\mapsto} #2}
\newcommand{\sublist}[1]{[#1]}
\newcommand{\subst}[2]{#1/#2}
\newcommand{\ssubst}[2]{\sublist{\subst{#1}{#2}}}
\newcommand{\substof}[1]{\sigma_{\!#1}}
\newcommand{\linkingof}[1]{\sim_{#1}}
\newcommand{\mapof}[1]{\lfloor{#1}\rfloor}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\quand}{\quad\mbox{and}\quad}

\newcommand{\qquand}{\qquad\mbox{and}\qquad}

@@

\begin{figure}
  \begin{center}
    \framebox{%
      $
      \begin{array}{@{}c@{}}
        \dbox{%
          $
          \begin{array}{c}
            \vlinf{\faD}{}{\Scons{\forall x.\ttt}}{\Scons\ttt}
            \qquad
            \vlinf{\aiD}{}{\Scons{a\vlor\cneg a}}{\Scons\ttt}
            \qquad
            \vlinf{\tttD}{}{\Scons{A\vlan\ttt}}{\Scons A}
            \\ \\[-1ex]
            \vlinf{\sw}{}{\Scons{(A\vlan B)\vlor C}}{\Scons{A\vlan (B\vlor C)}}
            \qquad
            \vlinf{\mix}{}{\Scons{A\vlor B}}{\Scons{A\vlan B}}            
            \\ \\[-1ex]
            \vlinf{\exD}{}{\Scons{\exists x.A}}{\Scons{A\ssubst x t}}
            \qquad
            \vlinf{\fequ}{\text{(where $A\fequ B$)}}{\Scons{A}}{\Scons{B}}
          \end{array}
          $
        }
        \\ \\[-1ex]
        \vlinf{\mfaD}{}{\Scons{\forall x.(A\vlor B)}}{\Scons{(\forall x.A)\vlor(\forall x.B)}}
        \qquad
        \vlinf{\mexD}{}{\Scons{\exists x.(A\vlor B)}}{\Scons{(\exists x.A)\vlor(\exists x.B)}}        
        \\ \\[-1ex]
        \vlinf{\me}{}{\Scons{(A\vlor B)\vlan(C\vlor D)}}{\Scons{(A\vlan C)\vlor(B\vlan D)}}
        \\ \\[-1ex]
        \vlinf{\wrD}{}{\Scons{A\vlor B}}{\Scons{A}}
        %%\qquad
        %%\vlinf{\wlD}{}{\Scons{B\vlor A}}{\Scons{A}}
        \qquad
        \vlinf{\acD}{}{\Scons{a}}{\Scons{a\vlor a}}
      \end{array}
      $
    }
  \end{center}
  \caption{Deep inference systems $\FOKS$ (all rules) and $\FOMLS$ (rules in the dashed box)}
  \label{fig:KS1}
\end{figure}

@@

\begin{thm}\label{thm:decomposition}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there are formulas $A_1,\ldots,A_5$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{A}{
        \vlde{\set{\acD}}{}{A_1}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A_2}{
            \vlde{\set{\exD}}{}{A_3}{
              \vlde{\set{\sw,\mix,\fequ}}{}{A_4}{
                \vlde{\set{\faD,\aiD,\tttD}}{}{A_5}{
                  \vlhy{\ttt}}}}}}}}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:decompositionA}
  For every derivation $\vlder{\FOKS}{\Deri}{A}{\ttt}$ there is a formula $A_1$, such that there is a derivation:
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{A}{
        \vlde{\FOMLS}{}{A_1}{
          \vlhy{\ttt}}}}
  \end{equation*}
\end{thm}

@@

\begin{thm}\label{thm:CP-DI}
  Let $A$ be a rectified formula, and assume we have a combinatorial proof $\phi\colon\gC\to\graphof A$. Then there is a derivation
  \begin{equation}
    \label{eq:decom}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri_2}{A}{
        \vlde{\FOMLS}{\Deri_1}{A'}{
          \vlhy{\ttt}}}}
  \end{equation}
  for some $A'$ such that $\graphof{\rectif{A'}}=\gC$. Conversely,
  whenever we have a derivation as in~\eqref{eq:decom} above, then
  there is a combinatorial proof $\phi\colon\gC\to\graphof A$ such that $\gC=\graphof{\rectif{A'}}$.
\end{thm}

@@

\begin{lemma}
  \label{lem:me}
  The rules $\me,\mfaD,\mexD$ are derivable in $\set{\wrD,\cD,\fequ}$. 
\end{lemma}

@@

\begin{lemma}
  Let $A$ and $B$ be formulas. Then
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{}{B}{
        \vlhy{A}}}
    \qquad
    \iff
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}
  \end{equation*}
\end{lemma}

@@

%All statements in this section are proved by rule permutation arguments.
\begin{thm}\label{thm:LK1-decompose}
  Let $\Gamma$ be a sequent. If\/ $\sqn\Gamma$ is provable in $\FOLK$ (as depicted on the left below) then there is a sequent
  $\Gamma'$ such that there is a derivation as shown on the right below:\vadjust{\vskip-2ex}
  \begin{equation*}
    \vlderivation{
        \vltrl{}{\FOLK}{\Deri}{\sqns{\Gamma}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}
    \qquad
    \Longrightarrow
    \qquad
    \vlderivation{
      \vlde{\set{\wrD,\cD,\fequ}}{\Deri_2}{\sqns{\form{\Gamma}}}{
        \vltrl{}{\FOMLL}{\Deri_1}{\sqns{\Gamma'}}{
          \vlhy{\quad}}{
          \vlhy{}}{
          \vlhy{}}}}
  \end{equation*}
\end{thm}
\juihsuan{maybe we should take $\form{\Gamma}$ instead? (consider the proof
$\vlderivation{\vlin{\wk}{}{\sqn{\ttt, A}}{
                 \vlin{\ttt}{}{\sqn\ttt}{
                   \vlhy{}}}}$)}

\begin{proof}
  Note that the instances of $\wrD,\cD$ in $\Deri_2$ are deep, but inside sequent contexts.
  \todo{}
\end{proof}

@@

\begin{lemma}\label{lem:cw-decomposition}
  For every derivation
  $\vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{}{B}{
        \vlhy{A}}}$
  there are formulas $A'$ and $B'$ such that 
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\fequ}}{}{B}{
        \vlde{\set{\acD}}{}{B'}{
          \vlde{\set{\me,\mfaD,\mexD,\fequ}}{}{A'}{
            \vlhy{A}}}}}
  \end{equation*}
\end{lemma}

@@

In this section we establish the relation between ske bifibrations and derivations  in
$\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}$. However, 
if a derivation $\Deri$ contains instances of the rules $\mfaD$ and
$\mexD$ we can no longer naively define the rectification
$\rectif\Deri$ as in the previous section for $\FOMLS$, as these two
rules cannot be applied if premise and conclusion are rectified. For
this reason we define here rectified versions $\rectif\mfaD$ and
$\rectif\mexD$, show below:
\begin{equation*}
  \vlinf{\rectif\mfaD}{}{\Scons{\forall x.(Ax\vlor Bx)}}{\Scons{(\forall y.Ay)\vlor(\forall z.Bz)}}
  \qquad
  \vlinf{\rectif\mexD}{}{\Scons{\exists x.(Ax\vlor Bx)}}{\Scons{(\exists y.Ay)\vlor(\exists z.Bz)}}
\end{equation*}
Here, we use the notation $A\cdot$ for a formula $A$ with occurrences
of a placeholder $\cdot$ for a variable. Then $Ax$ stands for the
results of replacing that placeholder with $x$, and also indicating
that $x$ must not occur in $A\cdot$. Then $\forall x.Ax$ and $\forall
y.Ay$ are the same formula modulo renaming of the bound variable bound
by the outermost $\forall$-quantifier. In the rules $\rectif\mfaD$ and
$\rectif\mexD$ we also demand that the variables $x$, $y$, and $z$ do
not occur in the context $\Sconhole$.

Note that in an instance of $\rectif\mfaD$ or $\rectif\mexD$ (as shown
above), we can have $x=y$ or $x=z$, but not both if the premise is
rectified. If $x=y$ and $x=z$ we have $\mfaD$ and
$\mexD$ as special cases of $\rectif\mfaD$ and $\rectif\mexD$, respectively.

For a derivation $\Deri$ in
$\set{\wlD,\wrD,\acD,\me,\mfaD,\mexD,\fequ}$, we can now construct the
\bfit{recification} $\rectif\Deri$ by rectifying each line of $\Deri$, yielding a derivation in 
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$.

For each instance $\vlinf{\rr}{}{P}{Q}$ of an inference rule in
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can
define the \bfit{induced map} $\mapof{\rr}\colon\vgraphof
Q\to\vgraphof P$ which acts as the identity for
$\rr\in\set{\me,\fequ}$ and as the canonical injection for
$\rr\in\set{\wlD,\wrD}$. For $\rr=\acD$ it maps the vertices
corresponding to the two atoms in the premise to vertex of the
contracted atom in the conclusion, and for
$\rr\in\set{\rectif\mfaD,\rectif\mexD}$ it maps the two vertices
corresponding to the quantifiers in the premise to the one in the
conclusion (as acts as the identity on all other vertices). For a
derivation $\Deri$ in
$\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we can then define
the \bfit{induced map} $\mapof\Deri$ as the composition of the induced
maps of the rule instances in $\Deri$.

\begin{lemma}
  Let $\downsmash{\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}}$ be a derivation. Then there is a recified derivation
  $\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{\rectif
      B}{ \vlhy{\rectif A}}}$, such that the induced maps
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ and
  $\mapof{\rectif\Deri}\colon\graphof{\rectif A}\to\graphof{\rectif
    B}$ are equal up to the relabeling of the vertices.
\end{lemma}

\begin{proof}
  Immediate from the definition.
\end{proof}

@@

\begin{lemma}\label{lem:cw->skew}
  Let $\vlderivation{
    \vlde{\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\Deri}{B}{
      \vlhy{A}}}$ be a recified derivation. Then the induced map
  $\mapof\Deri\colon\graphof{A}\to\graphof{B}$ is a skew bifibration.
\end{lemma}

@@

\begin{proof}[Proof of Lemma~\ref{lem:cw->skew}]
  First, observe that for every inference rule
  $\rr\in\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}$ we
  the induced map $\mapof{\rr}\colon\vgraphof Q\to\vgraphof P$ defines
  a fibration on the corresponding binding graphs. Therefore, their
  composition $\mapof\Deri$ is a fibration.

  Now we construct for
  $\Deri$ its propositional encoding $\PE\Deri$ by
  translating every line into its proposional encoding. The instances
  of the rules $\rectif\mfaD$ and $\rectif\mexD$ are replaced two
  steps by:
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlor(\PE{(Ax)}\vlor\PE{(Bx))}}}{
        \vlin{\fequ}{}{\Scons{(y\vlor z)\vlor(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlor\PE{(Ay)})\vlor (z\vlor\PE{(Bz)})}}}}}
  \end{equation*}
  and
  \begin{equation*}
    \vlderivation{
      \vlin{\rectif\acD}{}{\Scons{x\vlan\PE{(Ax)}\vlor\PE{(Bx)}}}{
        \vlin{\me}{}{\Scons{(y\vlor z)\vlan(\PE{(Ay)}\vlor\PE{(Bz)})}}{
          \vlhy{\Scons{(y\vlan\PE{(Ay)})\vlor (z\vlan\PE{(Bz)})}}}}}
  \end{equation*}
  respectively, where $\rectif\acD$ is a $\acD$ that renames the
  variabes---the atom variable, as well as the first-order variable of
  the same name---as everything is recified, there is no ambiguity
  here.
  This gives us a derivation $\vlderivation{
    \vlde{\set{\wrD,\acD,\rectif\acD,\me,\PE\fequ}}{\PE\Deri}{\PE
      B}{ \vlhy{\PE A}}}$ such that  $\mapof{\PE\Deri}=\mapof\Deri$. It has been
  shown in~\cite{str:07:RTA} that $\mapof{\PE\Deri}$ is a skew
  fibration (see
  also~\cite{hughes:invar,str:ral:tableaux19,str:RR-9048}). Hence,
  $\mapof\Deri$ is a skew fibration.
\end{proof}
\subsection{From Skew Bifibrations to Contraction and Weakening}

\begin{lemma}\label{skew->cw}
  Let $\gA$ and $\gB$ be fographs, let $\phi\colon\gA\to\gB$ be a skew bifibration, and let $A$ and $B$
  be formulas with $\graphof A=\gA$ and $\graphof B=\gB$. Then there
  are derivations
  \begin{equation*}
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\rectif\mfaD,\rectif\mexD,\fequ}}{\rectif\Deri}{B}{
        \vlhy{A}}}
    \quand
    \vlderivation{
      \vlde{\set{\wrD,\acD,\me,\mfaD,\mexD,\fequ}}{\Deri}{B}{
        \vlhy{A\substof\phi}}}
  \end{equation*}
  such that $\mapof{\rectif\Deri}=\phi$ and $\rectif{\Deri}$ is a
  rectification of $\Deri$, and $\substof\phi$ is the substitution
  induced by $\phi$.
\end{lemma}

\begin{proof}
  By Lemma~\ref{lem:PE} we have $\gA=\graphof{\PE A}$ and
  $\gB=\graphof{\PE B}$.  Let $\vB'\subseteq\vB$ be the image of
  $\phi$, and let $\gB_1$ be the subgraph of $\gB$ induced by
  $\vB'$. Hence, we have two maps
  $\phi''\colon\gA\to\gB_1$ being a surjection and $\phi'\colon
  \gB_1\to\gB$ being an injection that reflects edges. Both, $\phi'$
  and $\phi''$ remain skew bifibrations. Let us first look at $\phi'$.
  Let $\PE B_1$ be the propositional formula obtained from
  $\PE B$ by removing all atoms that are not represented by vertices
  in $\vB'$. Then $\graphof{\PE B_1}=\gB_1$. 
  By \cite[Proposition~7.6.1]{str:07:RTA}, we have a derivation
  $\vlder{\set{\wrD,\fequ}}{\PE\Deri_1}{\PE B}{\PE B_1}$.  A
  subformula of $\PE B$ is called \emph{weak} if it has been
  introduced by a weakening. All subformulas of a weak formula are
  also weak. Two weak subformulas $B'$ and $B''$ of $\PE B$ form a
  \emph{weak pair} if $B\fequ\Scons{B'\vlor B''}$ for some context
  $\Sconhole$. We now can assume without loss of generality that
  whenever weak subformulas $B'$ and $B''$ for a weak pair, they have
  been introduced by the same instance of $\wlD$ or $\wrD$ in
  $\PE\Deri_1$.\footnote{If $\PE\Deri_1$ is not of that shape, it can
  brought into this form by simple rule permutations, and then
  collapsing several weakenings into a single one.}  Now observe that
  whenever a weakening in $\PE\Deri_1$ deletes an atom $x\in\VAR$, it
  must also delete all atoms in the scope of the corresponding
  quantifier, because $\phi$ is a fibration on the binding
  graph. Hence, each line in $\PE\Deri_1$ is the propositional
  encoding $\PE P$ of a first-order formula $P$. Thus we have a first
  order derivation $\vlder{\set{\wrD,\fequ}}{\Deri_1}{B}{B_1}$. we now
  have to show that each instance of $\wlD$ or $\wrD$ is indeed a
  correct application in first-order logic. For this we have to
  inspect the cases a weakening happens inside a subformula $x\vlor C$
  or $x\vlan C$ in $\PE\Deri_1$. There are the following cases:
  \begin{equation*}%\small
    \vlinf{}{}{S\cons{x\vlor D\vlor C}}{S\cons{x\vlor C}}
    \qquad
    \vlinf{}{}{S\cons{x\vlan (D\vlor C)}}{S\cons{x\vlan C}}
    \qquad
    \vlinf{}{}{S\cons{(x\vlor D)\vlan C}}{S\cons{x\vlan C}}
  \end{equation*}
  In the first case the weakening happens inside the scope of a
  $\forall$-quantifier, and in the second case inside the scope of a
  $\exists$-quantifier. Both cases are unproblematic in first-order
  logic. However, in the third case an $\exists$-quantifier would be
  transformed into an $\forall$-quantifier. But as $\phi$ has to
  preserve existentials, this third case cannot occur, and therefore
  $\Deri_1$ is a valid first-order derivation in $\set{\wD,\fequ}$.
